---
title: "Pro Skills Group Project"
author: "Group 1"
format: html
editor: visual
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(corrr)
library(ggplot2)
library(tidyverse)
library(gt)
library(gridExtra)
library(MASS)
library(patchwork)
library(moderndive)
library(knitr)
library(kableExtra)
setwd("/Users/mattparker/Downloads")
data <- read.csv("project1-growth-wt.csv")
```

## Exploratory Analysis

```{r}
#| echo: false
#| fig-cap: Relationship between child weight at 24 months and weight at 1 month by sex
#| label: fig-scat-wt24-wt1
#| fig-align: center
#| fig-width: 4.5
#| fig-height: 3.5
ggplot(data, aes(x = Wt1, y = Wt24, color = Sex)) +
  geom_point() +
  labs(x = "Weight at 1 month (kg)", y = "Weight at 24 months (kg)", color = "Sex")
```

@fig-scat-wt24-wt1 shows the relationship between the child's weight at 24 months against the child's weight at 1 month. From @fig-scat-wt24-wt1 it appears that there is a positive, moderate to strong linear relationship between a child's weight at 24 months and weight at 1 month. Moreover, with the exception of a few points there appears to be a distinction by gender with female children having lower weight and male children having a higher weight.

```{r}
#| echo: false
#| warning: false
cors.male <- data |> filter(,Sex=="Male") |> correlate(,diagonal = 1)
cors.female <- data |> filter(,Sex=="Female") |> correlate(,diagonal = 1)
```

```{r}
#| echo: false
#| warning: false
#| tbl-cap: Correlations between numerical variables by sex
#| tbl-subcap: 
#|  - Correlations for numerical variables for male children
#|  - Correlations for numerical variables for female children
#| layout-ncol: 2
#| tbl-cap-location: bottom
#| label: tbl-corrs

# Male table
male_table <- cors.male |> 
  gt() |> 
  fmt_number(decimals = 2) |> 
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels())

# Female table
female_table <- cors.female |> 
  gt() |> 
  fmt_number(decimals = 2) |> 
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels())

# Print tables using layout-ncol: 2 to automatically place them side by side
male_table
female_table
```

@tbl-corrs shows the correlation between numerical variables for male and female children. From @tbl-corrs-1 we can see that for males Wt1 and Wt24 are moderately, positivly correlated, whereas Solids against Wt24 and Wt1 are weakly negatively correlated. @tbl-corrs-2 highlights that this is slightly different for female children as there appears to be a higher positive correlation between Wt24 and Wt1. Additionally, there appears to be a moderate negative correlation between Solids and Wt24 for female children and a weak-moderate correlation between Solids and Wt1. This suggests that there may be an interaction between Wt1 and Solids by Sex.

```{r}
#| echo: false
#| label: tbl-summary
#| tbl-cap:  Mean, median and standard deviation (sd) Wt24, Wt1 and Solids by sex.
data |>
   summarize('Wt24.Mean' = mean(Wt24),
             'Wt24.Median' = median(Wt24),
             'Wt24.sd' = sd(Wt24),
             'Wt1.Mean' = mean(Wt1),
             'Wt1.Median' = median(Wt1),
             'Wt1.sd' = sd(Wt1),
             'Solids.Mean' = mean(Solids),
             'Solids.Median' = median(Solids),
             'Solids.sd' = sd(Solids),
             .by = Sex) |>
 gt() |> 
  fmt_number(decimals=2) |>
  tab_spanner(
    label = "Wt24",
    columns = c(Wt24.Mean, Wt24.Median,Wt24.sd)
  ) |>
  tab_spanner(
    label = "Wt1",
    columns = c(Wt1.Mean, Wt1.Median,Wt1.sd)
  ) |>
    tab_spanner(
    label = "Solids",
    columns = c(Solids.Mean, Solids.Median, Solids.sd)
  ) |>
  cols_label(
    Wt24.Mean = html("Mean"),
    Wt24.Median = html("Median"),
    Wt24.sd = html("Std.Dev"),
    Wt1.Mean = html("Mean"),
    Wt1.Median = html("Median"),
    Wt1.sd = html("Std.Dev"),
    Solids.Mean = html("Mean"),
    Solids.Median = html("Median"),
    Solids.sd = html("Std.Dev")
  )
```

@tbl-summary shows the mean, median and standard deviation of Wt24, Wt1 and Solids by Sex. @tbl-summary highlights that there appears to be a slight difference between Wt24 by sex. There does not appear to be a difference in Wt1 and Solids by sex.
